---
title: "Orthology Results: Assessing the initial orthogroups (homogroups) 46 metazoans"
author: "Dean Mckeown"
date: "`r Sys.Date()`"
output: 
  html_document:
    output_file: "../results/homogroups.html"
    toc: true
    toc_float: TRUE
    toc_depth: 4
    number_sections: true
    theme: united
    highlight: tango
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "../results/")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{css, echo=FALSE}
.scroll-200 {
  max-height: 200px;
  overflow-y: auto;
  background-color: inherit;
}
```

___

## Preparation
We will begin by creating a variable that specifies where all workflow outputs are stored. Additionally we will source the script containing helper functions, and setting to variable some colors that we'll use throughout. 
### Inputs needed

/users/asebe/dmckeown/projects/crg-bcaortho/project-ioo_mz46/results/orthofinder_results/orthofinder_mcl/Orthogroups.tsv
  * **Pipeline generated, for any orthocaller** - which homogroups contain what sequences

/users/asebe/dmckeown/projects/crg-bcaortho/project-ioo_mz46/results/prefilter/initial/defline_info/*.csv
  * **Pipeline generated, for any orthocaller** - hmmsearch assignment for each sequence

/users/asebe/dmckeown/projects/crg-bcaortho/project-ioo_mz46/results/orthofinder_results/orthofinder_mcl/Orthogroups.GeneCount.tsv
  * **Pipeline generated, for any orthocaller** - just Orthogroups.tsv just totals instead of deflines

/users/asebe/xgraubove/genomes/annotation_functional/*_long.pep.pfamscan.csv
  * **Precalculated functional annotations**, needs gene id, PFAM accession for OG homogeneity assessment by cogeqc

### Inputs needed, but currently only available for orthofinder
/users/asebe/dmckeown/projects/crg-bcaortho/project-ioo_mz46/results/orthofinder_mcl/complete_dataset/Results_Inflation_2.5/Comparative_Genomics_Statistics/Statistics_PerSpecies.tsv
  * Basic stats of ortholog size etc by species

/users/asebe/dmckeown/projects/crg-bcaortho/project-ioo_mz46/results/orthofinder_mcl/complete_dataset/Results_Inflation_2.5/Comparative_Genomics_Statistics/Orthogroups_SpeciesOverlaps.tsv
  * matrix of all samples vs all - might need to hunt down OF code
  * the number of orthogroups shared between each species-pair as a square matrix.

```{r Preparation, results = FALSE}
suppressPackageStartupMessages(suppressWarnings(source('R/packages.R')))
suppressPackageStartupMessages(suppressWarnings(source('R/summary-functions.R')))
source('R/arcadia-color-gradients.R')

# INPUTS
# Specify where the workflow outputs are stored (relative paths unless indicated)

RESULT_DIR  <-  "../results/orthofinder_results/"
SAMPLESHEET <- "/users/asebe/dmckeown/projects/crg-bcaortho/project-ioo_mz46/samplesheet1.csv"

STATS_PER_SPP_TSV <- "../results/orthofinder_results/orthofinder_mcl/complete_dataset/Results_Inflation_2.5/Comparative_Genomics_Statistics/Statistics_PerSpecies.tsv"
GENE_COUNTS_TSV <- "../results/orthofinder_results/orthofinder_mcl/Orthogroups.GeneCount.tsv"

# OUTPUTS
# And where summarized results should be saved.
OUTPUT_DIR <- './results/'
dir.create(OUTPUT_DIR, showWarnings = F)
dir.create(paste0(OUTPUT_DIR, 'orthofinder'), showWarnings = F)

# OTHER PARAMETERS
# Set colours to plot supergroups (taxonomy field in samplesheet1)

SUPERGROUP_COLS <- c(
  Choanoflagellata = "#BEBADA",
  Cnidaria = "#FB8072",
  Ctenophora = "#80B1D3",
  Deuterostomia = "#8DD3C7",
  Filasterea = "#FDB462",
  Placozoa = "#B3DE69",
  Porifera = "#FCCDE5",
  Protostomia = "#FFFFB3",
  Teretosporea = "#D9D9D9"
)

```

___
    
___

## Gene Family Summaries {.tabset}
Now, let's go ahead and summarize the results from OrthoFinder. To start, this will just include some high-level summaries - number of species per gene family, gene copy number per gene family, etc.

```{r Gene Family Summary Stats, fig.asp = .95}
# Summarize gene family content, both with respect to number of species 
# included in each gene family, and the per-species mean copy number
summ_genefam_composition(results_dir = RESULT_DIR, show_plots = T, out_dir = paste0(OUTPUT_DIR, 'orthofinder/'))

# And get orthogroup summary statistics per species
per_spp_of_stats <-
  get_per_spp_ofinder_stats(tree_fpath = paste0(RESULT_DIR, 'speciesrax/species_trees/inferred_species_tree.newick'),
                        samplesheet_fpath = SAMPLESHEET,
                        results_dir = RESULT_DIR, grp_name = 'Supergroup',
                        out_dir = paste0(OUTPUT_DIR, 'orthofinder/'),
                        tip_grp_cols = SUPERGROUP_COLS, 
                        count_cols = arcadia_cividis,
                        prop_cols = arcadia_viridis,
                        og_stats_perspp_fpath = STATS_PER_SPP_TSV)

# Lastly, get the per-species orthogroup counts/copy numbers
og_counts <- get_per_spp_og_counts(results_dir = RESULT_DIR, gene_count_tsv = GENE_COUNTS_TSV)
```

___

## Homogroup Homogeneity
